---
title: "Evaluating signatures using their original models"
author: 
- name: Xutao Wang
  affiliation: 
  - Department of Biostatistics, Boston University, Boston, MA
  email: xutaow@bu.edu

package: TBSignatureProfiler
output: 
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{"Evaluating signatures using their original models"}
  %\VignetteEncoding{UTF-8}  
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

# Introduction
This document provides some examples, which helps you to evaluate certain signature's performance using its original models. Currently, the package has incoporated the original methods for the following gene signatures. The specifc genes within each biomarker can be found at `TBsignatures`. Users should choose signatures from the below list when using the original model feature.

```{r}
library(TBSignatureProfiler)
signatureOriginalModel <- c("Anderson_42", "Anderson_OD_51", "Kaforou_27",
                            "Kaforou_OD_44", "Kaforou_OD_53", "Sweeney_OD_3",
                            "Maertzdorf_4", "Maertzdorf_15", "LauxdaCosta_OD_3",
                            "Verhagen_10", "Jacobsen_3", "Sambarey_HIV_10",
                            "Leong_24", "Berry_OD_86", "Berry_393",
                            "Bloom_OD_144", "Suliman_RISK_4", "Zak_RISK_16",
                            "Leong_RISK_29", "Zhao_NANO_6")
```

# Evaluation
In this tutorial, we will work with HIV and Tuberculosis (TB) gene expression data in a `SummarizedExperiment` format. First of all, we evaluate the performance of all available TB gene signatures whose original models have been included in the package. 

```{r, message=FALSE}
## HIV/TB gene expression data, included in the package
hivtb_data <- TB_hiv
out <- evaluateOriginalModel(input = hivtb_data, useAssay = "counts")
```

Users can also evaluate selected gene signatures based on their preference.

```{r, message=FALSE}
outSub <- evaluateOriginalModel(input = hivtb_data, useAssay = "counts",
                                geneSignaturesName = c("Anderson_42",
                                                       "Sweeney_OD_3",
                                                       "Verhagen_10",
                                                       "Zak_RISK_16"))
```

The returned object is also of the `SummarizedExperiment`. The scores will be returned as a part of the colData with column names: "Name_Of_Signaure_OriginalModel". The structure of the returned object is the same as the one given by `runTBsigProfiler`. Users can now follow the guide from TBSignatureProfiler main [package vignette](http://www.bioconductor.org/packages/release/bioc/vignettes/TBSignatureProfiler/inst/doc/tbspVignette.html) for downstream analysis.

